UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-05-02 ms/main.tex
shortened UMAP-tsne comparison

\subsubsection*{Comparing t-SNE and UMAP}
To compare UMAP to t-SNE and assess the role of convergence, we ran UMAP, Python's sklearn implementation of t-SNE, and a hybrid approach where the early exaggeration phase of t-SNE (typically the first 250 iterations) has been replaced by a full UMAP run (600 epochs for the 1KGP data and 230 epochs for the HRS and UKBB data). The three approaches produce similar results in smaller datasets such as the 1KGP and HRS, with the main differences being in run-time and some qualitative factors such as clusters being more interconnected when using UMAP (figures~\ref{fig:supp_tsne_umap_compare_1kgp},~\ref{fig:supp_tsne_umap_compare_hrs}). When applied to the UKBB, UMAP provides a more compelling visualization of the data compared to both versions of t-SNE. Interestingly, the hybrid approach produces a better value for the t-SNE metric than the default t-SNE approach (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}), suggesting that UMAP can help standard t-SNE implementation to reach the t-SNE objective. The standard t-SNE initialization generated a large number of arbitrary clusters, while initializing t-SNE with UMAP preserved much of the large scale structure created by the UMAP embedding but lost much fine-scale resolution and the relationship among clusters (figure~\ref{fig:supp_tsne_umap_compare_ukbb}). An animation of t-SNE iterations starting from the UMAP embedding can be seen in the supplementary files. Given these results, we recommend UMAP over t-SNE for large and diverse genomic datasets.

>>>>

\subsubsection*{Comparing t-SNE and UMAP}

Identifying the best dimension reduction technique is to an extent subjective. However, we did find that using UMAP to preprocess the UKBB data allowed a standard t-SNE implementation to find representations that were objectively better (according to the t-SNE metric) than the default t-SNE implementation (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}). Yet, these representations were much less detailed than the UMAP embedding provided as a starting point (figure~\ref{fig:supp_tsne_umap_compare_ukbb}.  Given these results, we recommend UMAP over t-SNE for large and diverse genomic datasets.

%To compare UMAP to t-SNE and assess the role of convergence, we ran UMAP, Python's sklearn implementation of t-SNE, and a hybrid approach where the early exaggeration phase of t-SNE (typically the first 250 iterations) has been replaced by a full UMAP run (600 epochs for the 1KGP data and 230 epochs for the HRS and UKBB data). The three approaches produce similar results in smaller datasets such as the 1KGP and HRS, with the main differences being in run-time and some qualitative factors such as clusters being more interconnected when using UMAP (figures~\ref{fig:supp_tsne_umap_compare_1kgp},~\ref{fig:supp_tsne_umap_compare_hrs}). When applied to the UKBB, UMAP provides a more compelling visualization of the data compared to both versions of t-SNE. Interestingly, the hybrid approach produces a better value for the t-SNE metric than the default t-SNE approach (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}), suggesting that UMAP can help standard t-SNE implementation to reach the t-SNE objective. The standard t-SNE initialization generated a large number of arbitrary clusters, while initializing t-SNE with UMAP preserved much of the large scale structure created by the UMAP embedding but lost much fine-scale resolution and the relationship among clusters (figure~\ref{fig:supp_tsne_umap_compare_ukbb}). An animation of t-SNE iterations starting from the UMAP embedding can be seen in the supplementary files. Given these results, we recommend UMAP over t-SNE for large and diverse genomic datasets.
